Agda Meets Accelerate

### 2012-08-30 paper/main.tex
small fixes

To see the Agda embedding in action, we translate the
dot product example from Section~\ref{sec:accelerate} to Agda.
\begin{code}
dotp : {E : Element} {{p : IsNumeric E}} {n : Nat}
     -> PreVector n E -> PreVector n E -> Scalar E
dotp{E} xs ys = 
  let xs' = use xs
      ys' = use ys
  in
  fold _+_ ("0" ::: E) (zipWith _*_ xs' ys')
\end{code}
Unlike the Accelerate code, this function is polymorphic with respect
to the array element type, provided it is numeric.\footnote{In Agda, arguments in double curly braces are \emph{instance arguments}~\cite{DevriesePiessens2011}. We use them much like type class constraints are used in Haskell.} It furthermore
takes the length $n$ as a parameter thus ensuring that the two input
vectors have the same size. The \texttt{PreVector} type of the
arguments corresponds to the plain \texttt{Vector} type in Accelerate,
whereas the result type \texttt{Scalar E} corresponds to \texttt{Acc
  (Scalar E)}---a piece of abstract syntax.

@@

In any case, providing such information would be done by including an
additional argument that holds a suitable proof object, as in
\begin{verbatim}
fold : {E}{sh}{n} -> (f : Exp E -> Exp E -> Exp E) -> (e : Exp E)
     -> Array (sh :< n >) E -> IsMonoid f e -> Array sh E
\end{verbatim}
where
\begin{verbatim}
IsMonoid : {E} -> (f : Exp E -> Exp E -> Exp E) -> (e : Exp E) -> Set
IsMonoid f e = ( IsAssociative f , IsUnit f e)
\end{verbatim}

@@

For a related problem, consider an implementation of the \texttt{filter}
operation that takes a predicate and a source array and returns an
array that only contains the elements of the source array fulfilling
the predicate.  First of all, filtering only makes sense for
one-dimensional arrays, that is, for vectors. To see the second catch,
let's try to write down a dependent type signature for \texttt{filter}.
\begin{verbatim}
filter : {n m : Nat}{E : Element}
       -> Vector n E -> (Exp E -> Exp Bool) -> Vector m E
\end{verbatim}
The problem is that the size of the result
cannot be determined statically. In fact, the only thing we know about
\texttt{m} is that it must be less than or equal to
\texttt{n}. However, we cannot prove this from the code because of a staging restriction. The Accelerate implementation computes the length of the result only when the generated GPU code is executed. The function \texttt{Exp E -> Exp Bool} maps abstract syntax to abstract syntax; it does not directly implement a Boolean predicate. Hence, we cannot use the predicate in the result type of \texttt{filter} to more accurately constrain the size of the resulting vector --- unless we include an evaluator for Accelerate expressions.

@@

However, an alternative encoding of arrays can be
used which is compatible with filtering of elements. The idea of this
encoding is to keep all elements but mark those which are no longer
present because they have been filtered out.
There are several ways of implementing this idea. The simplest
approach is to wrap each element in a maybe type or
pair up each element with a boolean flag that indicates
its presence.\footnote{A \texttt{Maybe} type is on one hand the better option,
  but it has to be coded without using pattern matching.}
\begin{verbatim}
FVector : Nat -> Element -> Set
FVector n E = Vector n (Pair Bool E)
\end{verbatim}

Now filtering becomes quite simple because the length of
the\texttt{FVector} does not change. Furthermore, filtering could be
extended to multi-dimensional arrays, although it is not clear if an
array that includes non-existing elements is a sensible notion.
\begin{verbatim}
filterF : {n : Nat}{E : Element}
        -> (Exp E -> Exp Bool) -> FVector n E -> FVector n E
filterF {n}{E} pred vec =
  map g vec
  where
  g : Exp (Pair Bool E) -> Exp (Pair Bool E)
  g bx = pair ((fst bx) && p (snd bx)) x
\end{verbatim}
However, mapping becomes more complicated because it either has to
materialize a dummy result for each absent element in the argument
vector or apply the function to absent elements, too.\footnote{This
  complication could be avoided with the \texttt{Maybe} type.}
\begin{verbatim}
mapF : {n : Nat}{E F : Element}
     -> Exp F -> (Exp E -> Exp F) -> FVector n E -> FVector n F
mapF {n}{E}{F} defaultF f vec =
  map g vec
  where
  g : Exp (Pair Bool E) -> Exp (Pair Bool F)
  g bx = if (fst bx) then (pair (fst bx) (f (snd bx)))
                     else (pair (fst bx) defaultF)
\end{verbatim}
On the positive side, some operations can get rid of the absent
elements. In particular, a fold operation which reduces a filtered
vector with a monoid returns a single value. In Accelerate, such a value has type
\texttt{Scalar}, which is a synonym for an array of dimension $0$.
\begin{verbatim}
foldF : {n : Nat}{E : Element}
      -> (Exp E -> Exp E -> Exp E) -> Exp E
      -> FVector n E -> Scalar E
foldF f e vec =
  fold f e (map (\ bx -> if (fst bx) then (snd bx) else e) vec)
\end{verbatim}
Other operations like \texttt{fold1} and the scan operations present
in Accelerate can also be lifted to this representation, but they
retain a notion of absent elements and do not allow to revert to a
non-filtered representation.

@@

Here is the Agda adapter for \texttt{use}:
\begin{verbatim}
use : {sh : Shape}{E : Element} -> PreArray sh E -> Array sh E
use {sh}{E} (PA y) = Ar (useHs (EltDict E) y)
\end{verbatim}
It makes use of two wrapper types. \texttt{PreArray} wraps a
one-dimensional Haskell array using the constant \texttt{HsDIM1} (the
\texttt{DIM1} type shown in Section~\ref{sec:accelerate} imported from
Haskell via FFI) and the function \texttt{EltType} (not shown), which
interprets a value of type \texttt{Element} as a Haskell type. The
latter types are also imported via FFI.
\begin{code}
data PreArray (E : Element) : Shape -> Set where
  PA : {sh : Shape} -> HsArray HsDIM1 (EltType E) -> PreArray sh E
\end{code}
The \texttt{Array} type wraps an AST reference for an Accelerate
array, where \texttt{Acc} and \texttt{AccArray} are types imported
from Haskell.
\begin{code}
data Array (E : Element) : Shape -> Set where
  Ar : {sh : Shape} -> Acc (AccArray HsDIM1 (EltType E)) -> Array sh E
\end{code}
The \texttt{EltDict} function translates a value 
\texttt{(E : Element)} into a Haskell expression that evaluates to a
dictionary for the Haskell type of \texttt{E} for the Haskell type
class \texttt{Elt}. Such a dictionary is passed, whenever that
corresponding Haskell function has type class constraints.
\begin{code}
EltDict : (E : Element) -> HsEltDict (EltType E)
\end{code}

@@

As an example for a function that requires more work on either side,
consider the \texttt{fold} operation. 
\begin{code}
fold : {E}{sh}{n}
     -> (Exp E -> Exp E -> Exp E)
     -> Exp E
     -> Array (sh :< n >) E
     -> Array sh E
fold {E}{sh}{n} f (Ex e) (Ar a) =
  Ar (foldHs (EltDict E) (toHsInt (size sh)) (toHsInt n)
             (unwrap2 f) e a)
\end{code}
As values of type \texttt{Exp} also need a wrapper type in Agda (it is
not possible to import type constructors via the FFI), there is some
unwrapping going on for the \texttt{e} and \texttt{f}
arguments. The implementation of \texttt{fold} just calls the
\texttt{foldHs} function and encodes the information about the shape
in two integer arguments. Here, \texttt{size sh} is the size of the
result and \texttt{n} is the size of the dimension that is folded. As
these values are initially available as Agda natural numbers, they need to
be converted to Haskell numbers using the function \texttt{toHsInt}.

>>>>

To see the Agda embedding in action, we translate the
dot product example from Section~\ref{sec:accelerate} to Agda.
\begin{code}
dotp : forall {E : Element} {{p : IsNumeric E}} {n : Nat}
     -> PreVector n E -> PreVector n E -> Scalar E
dotp{E} xs ys = 
  let xs' = use xs
      ys' = use ys
  in
  fold _+_ ("0" ::: E) (zipWith _*_ xs' ys')
\end{code}
Unlike the Accelerate code, this function is polymorphic with respect
to the array element type, provided it is numeric.\footnote{In Agda, arguments in double curly braces are \emph{instance arguments}~\cite{DevriesePiessens2011}. We use them much like type class constraints are used in Haskell.} It furthermore
takes the length $n$ as a parameter thus ensuring that the two input
vectors have the same size. The \texttt{PreVector} type of the
arguments corresponds to the plain \texttt{Vector} type in Accelerate,
whereas the result type \texttt{Scalar E} corresponds to \texttt{Acc
  (Scalar E)}---a piece of abstract syntax.

@@

In any case, providing such information would be done by including an
additional argument that holds a suitable proof object, as in
\begin{verbatim}
fold : forall {E}{sh}{n} -> (f : Exp E -> Exp E -> Exp E) -> (e : Exp E)
     -> Array (sh :< n >) E -> IsMonoid f e -> Array sh E
\end{verbatim}
where
\begin{verbatim}
IsMonoid : forall {E} -> (f : Exp E -> Exp E -> Exp E) -> (e : Exp E) -> Set
IsMonoid f e = ( IsAssociative f , IsUnit f e)
\end{verbatim}

@@

For a related problem, consider an implementation of the \texttt{filter}
operation that takes a predicate and a source array and returns an
array that only contains the elements of the source array fulfilling
the predicate.  First of all, filtering only makes sense for
one-dimensional arrays, that is, for vectors. To see the second catch,
let's try to write down a dependent type signature for \texttt{filter}.
\begin{verbatim}
filter : forall {n m : Nat}{E : Element}
       -> Vector n E -> (Exp E -> Exp Bool) -> Vector m E
\end{verbatim}
The problem is that the size of the result
cannot be determined statically. In fact, the only thing we know about
\texttt{m} is that it must be less than or equal to
\texttt{n}. However, we cannot prove this from the code because of a staging restriction. The Accelerate implementation computes the length of the result only when the generated GPU code is executed. The function \texttt{Exp E -> Exp Bool} maps abstract syntax to abstract syntax; it does not directly implement a Boolean predicate. Hence, we cannot use the predicate in the result type of \texttt{filter} to more accurately constrain the size of the resulting vector --- unless we include an evaluator for Accelerate expressions.

@@

However, an alternative encoding of arrays can be
used which is compatible with filtering of elements. The idea of this
encoding is to keep all elements but mark those which are no longer
present because they have been filtered out.
There are several ways of implementing this idea. The simplest
approach is to pair up each element with a boolean flag that indicates
its presence.\footnote{The \texttt{Maybe} type is not an option
  because Accelerate does not support \texttt{Maybe} types as array elements.}
\begin{verbatim}
FVector : Nat -> Element -> Set
FVector n E = Vector n (Pair Bool E)
\end{verbatim}

Now filtering becomes quite simple because the length of
the\texttt{FVector} does not change. Furthermore, filtering could be
extended to multi-dimensional arrays, although it is not clear if an
array that includes non-existing elements is a sensible notion.
\begin{verbatim}
filterF : forall {n : Nat}{E : Element}
        -> (Exp E -> Exp Bool) -> FVector n E -> FVector n E
filterF {n}{E} pred vec =
  map g vec
  where
  g : Exp (Pair Bool E) -> Exp (Pair Bool E)
  g bx = pair ((fst bx) && p (snd bx)) x
\end{verbatim}
However, mapping becomes more complicated because it either has to
materialize a dummy result for each absent element in the argument
vector or apply the function to absent elements, too.\footnote{This
  complication could be avoided with the \texttt{Maybe} type.}
\begin{verbatim}
mapF : forall {n : Nat}{E F : Element}
     -> Exp F -> (Exp E -> Exp F) -> FVector n E -> FVector n F
mapF {n}{E}{F} defaultF f vec =
  map g vec
  where
  g : Exp (Pair Bool E) -> Exp (Pair Bool F)
  g bx = if (fst bx) then (pair (fst bx) (f (snd bx)))
                     else (pair (fst bx) defaultF)
\end{verbatim}
On the positive side, some operations can get rid of the absent
elements. In particular, a fold operation which reduces a filtered
vector with a monoid returns a single value. In Accelerate, such a value has type
\texttt{Scalar}, which is a synonym for an array of dimension $0$.
\begin{verbatim}
foldF : forall {n : Nat}{E : Element}
      -> (Exp E -> Exp E -> Exp E) -> Exp E
      -> FVector n E -> Scalar E
foldF f e vec =
  fold f e (map (\ bx -> if (fst bx) then (snd bx) else e) vec)
\end{verbatim}
Other operations like \texttt{fold1} and the scan operations present
in Accelerate can also be lifted to this representation, but they
retain a notion of absent elements and do not allow to revert to a
non-filtered representation.

@@

Here is the Agda adapter for \texttt{use}:
\begin{verbatim}
use : forall {sh : Shape}{E : Element} -> PreArray sh E -> Array sh E
use {sh}{E} (PA y) = Ar (useHs (EltDict E) y)
\end{verbatim}
It makes use of two wrapper types. \texttt{PreArray} wraps a
one-dimensional Haskell array using the constant \texttt{HsDIM1} (the
\texttt{DIM1} type shown in Section~\ref{sec:accelerate} imported from
Haskell via FFI) and the function \texttt{EltType} (not shown), which
interprets a value of type \texttt{Element} as a Haskell type. The
latter types are also imported via FFI.
\begin{code}
data PreArray (E : Element) : Shape -> Set where
  PA : {sh : Shape} -> HsArray HsDIM1 (EltType E) -> PreArray sh E
\end{code}
The \texttt{Array} type wraps an AST reference for an Accelerate
array, where \texttt{Acc} and \texttt{AccArray} are types imported
from Haskell.
\begin{code}
data Array (E : Element) : Shape -> Set where
  Ar : {sh : Shape} -> Acc (AccArray HsDIM1 (EltType E)) -> Array sh E
\end{code}
The \texttt{EltDict} function translates a value 
\texttt{(E : Element)} into a Haskell expression that evaluates to a
dictionary for the Haskell type of \texttt{E} for the Haskell type
class \texttt{Elt}. Such a dictionary is passed, whenever that
corresponding Haskell function has type class constraints.
\begin{code}
EltDict : (E : Element) -> HsEltDict (EltType E)
\end{code}

@@

As an example for a function that requires more work on either side,
consider the \texttt{fold} operation. 
\begin{code}
fold : forall {E}{sh}{n}
     -> (Exp E -> Exp E -> Exp E)
     -> Exp E
     -> Array (sh :< n >) E
     -> Array sh E
fold {E}{sh}{n} f (Ex e) (Ar a) =
  Ar (foldHs (EltDict E) (toHsInt (size sh)) (toHsInt n)
             (unwrap2 f) e a)
\end{code}
As values of type \texttt{Exp} also need a wrapper type in Agda (it is
not possible to import type constructors via the FFI), there is some
unwrapping going on for the \texttt{e} and \texttt{f}
arguments. The implementation of \texttt{fold} just calls the
\texttt{foldHs} function and encodes the information about the shape
in two integer arguments. Here, \texttt{size sh} is the size of the
result and \texttt{n} is the size of the dimension that is folded. As
these values are initially available as Agda natural numbers, they need to
be converted to Haskell numbers using the function \texttt{toHsInt}.
